---
title: "Exploratory Data Analysis"
author: "JJC"
subtitle: Motor Trends Project
output:
  pdf_document:
    highlight: tango
    number_sections: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)

library(data.table)
```

# Reflect on the Research Question 
1.  "Is an automatic or manual transmission better for MPG"
2.  "How different is the MPG between automatic and manual transmission?"

In question 1, there is a comparison between a categorical variable (transmission) and the effect on response variable (MPG).  "better" in this context means higher miles per gallon.

Question 2 calls for a quantitative comparision between the two transmission categories.  

## Data Narrative

### Univariable numerical summaries
```{r}
t_count <- 1
```
_Table `r t_count`. Types of variables._  
Inspect first 6 observations to determine types of variables and their values.
```{r univariate, echo = FALSE}
apply(mtcars, 2, class)
```

* Convert categorical variables to factor.
```{r}
t_count <- t_count + 1
```
_Table `r t_count`. Quartile summaries_  
Quartile summaries for _mpg_:
```{r cast am}
data(mtcars)
mtcars <- data.table(mtcars)
mtcars$am_f <- factor(mtcars$am,
                      levels = c(0, 1),
                      labels = c("auto", "manual"))
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)
summary(mtcars$mpg)
```

Counts for each transmission type:
```{r}
am_mx <- t( as.matrix(table(mtcars$am)) )
colnames(am_mx) <- c("auto", "manual")
rownames(am_mx) <- ""
am_mx
```

### Distribution of Response Variable
Linear regression requires that the mean of the response is normal.
```{r mpg histogram, fig.align="left"}
# ref: Kabacoff, R in Action, pg 125

x <- mtcars$mpg
h <- hist(x, breaks = 12,
        col = "red",
        xlab = "mpg", 
        main = "Figure 1. Histogram of mpg with Normal curve")
xfit <- seq(min(x), max(x), length = 40)
yfit <- dnorm(xfit, mean = mean(x), sd = sd(x))
yfit <- yfit * diff(h$mids[1:2]) * length(x)
lines(xfit, yfit, col = "blue", lwd = 2)
box()

```

By superimposing a normal curve, figure 1 shows that the distribution of mpg values is approximately normal. 

### Bivariate associations

__Numeric Variables__  
1. Correlation with respect to _mpg_
```{r correlations}
x <- mtcars[, c("disp", "hp", "drat", "wt", "qsec")]
y <- mtcars[, c("mpg")]
cor_mx <- cor(x, y)
colnames(cor_mx) <- "mpg"
cor_mx
```
Three variables (_disp_, _hp_ and _wt_) have high negative correlation with _mpg_.  So the more the powerful engine and the heavier the car leads to lower fuel efficiency. _qsec_ has only moderate positive correlation with mpg and so, it does not appear in figure 2. The canon for high correlation here is an absolute value greater than 0.75. 

2. Scatterplots

```{r}
num_measures <-  mtcars[, c("disp", "hp", "drat", "wt", "mpg", "am")]

## put histograms on the diagonal
panel.hist <- function(x, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(usr[1:2], 0, 1.5) )
    h <- hist(x, plot = FALSE)
    breaks <- h$breaks; nB <- length(breaks)
    y <- h$counts; y <- y/max(y)
    rect(breaks[-nB], 0, breaks[-1], y, col = "blue", ...)
}

pairs( ~ disp + hp + drat + wt + mpg, data = mtcars,
       main = "Figure 2. Motor Trends Measures - 2 transmission types",
      pch = 21, bg = c("red", "green3")[unclass(mtcars$am)],
      cex = 1.5, 
      upper.panel = NULL, 
      diag.panel = panel.hist, cex.labels = 2)
legend("topright", legend = c("manual", "auto"),
       col = c("red", "green3"),
       text.col = c("red", "green3"), pch = c(16, 16),
       merge = FALSE, bg = "gray90")
```
* In figure 2, the lowest row confirms the correlation between _mpg_ and the four numerical explanatory variables.  
* Automatic transmission (green points) tends to have miles per gallon except for the _drat_ measure.
* Note that here is high correlation between the explanatory variables themselves.  So checking for multicollinearity in linear regression diagnostics is important.  
3. Correlation

```{r}
library(corrplot)

cor_mx <- cor(num_measures)
cor_mx
corrplot(cor_mx)

cor(mtcars$wt, mtcars$disp)
```
high correlation between wt and disp

__Categorical Variables__

Check transmission groups by wt and by disp

```{r}
par(mfrow = c(1, 2))
boxplot(wt ~ am_f, data = mtcars,
        xlab= "Transmission type",
        ylab = "weight 1000 lbs",
        main = "Figure 3  Weight by Transmission Groups")
boxplot(disp ~ am_f, data = mtcars,
        xlab= "Transmission type",
        ylab = "Displ cu.in",
        main = "Figure 4  Displ by Transmission Groups")

par(mfrow=c(1, 1))
```


Figure 3 orders each car model by fuel mileage for the two transmission groups.  Again, the _auto_ group has higher _mpg_.  However, there is crossover between the two groups where models have similar mileage values. 

```{r}
x <- mtcars[order(mtcars$mpg), ]
x$am <- factor(x$am, labels = c("manual", "auto"))

x$colour[x$am == "manual"] <- "red"
x$colour[x$am == "auto"] <- "green3"
dotchart(x$mpg,
         labels = rownames(x),
         cex = 0.7,
         groups = x$am,
         gcolor = "black",
         color = x$colour,
         pch = 19,
         main = "Figure 3, Car mileage for two transmission categories.",
         xlab = "Miles per Gallon"
)

```

Question 1 asks if one group is "better". This seeks to determine if the mean of one group is significantly higher than the other.  One solution is to perform a single-tail t-test.

### Assumptions for t-test
1. __Independent Sample__  
Differentiating by transmission category results in two independent samples since the manual group does not effect the automatic group.  
2. __Independent Observations__  
The selection of one car model does not effect the selection of another model.
3. __Normality Assumption__  
Figure 2 shows that the manual group has a normal distribution.  However,  the automatic group has right skew.  Also, it only contains 13 observations.  
```{r}
t_count <- 1
```
  
**Table `r t_count`. _mpg_ numeric summaries by transmission group.**
```{r mpg numerical summary, results='asis'}
library(kableExtra)
options(digits = 3)
mpg_mx <- matrix( c(mtcars[am == 0,
                           c(.N, mean(mpg), sd(mpg), min(mpg), max(mpg))]),
                  nrow = 1, ncol = 5)
mpg_mx <- rbind(mpg_mx, mtcars[am == 1, c(.N, mean(mpg), sd(mpg),
                                           min(mpg), max(mpg))] )

colnames(mpg_mx) <- c("n", "mean", "sd", "min", "max")
rownames(mpg_mx) <- c("manual", "auto")
mpg_mx %>%
  kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", full_width = F,
                                      font_size = 12))

```

However, Table `r t_count` shows that both groups are within 2.5 times standard deviation of the mean.  This suggests that a t-test is valid with a caveat about the distribution of the _am_ group.

For question 2, a linear model with a single categorical regressor variable would quantify the difference between the two transmission categories.  It would be worth comparing a model with only _am_ as regressor to a model with _am_ that accounts for all the other numeric variables. Accounting for the remaining categorical variables would require too many comparisons due to the number of levels involved. 

### mpg and binary categorical variables

t-test, Independent Samples
mpg by transmission category
```{r}
attach(mtcars)
mpg_a <- mtcars$mpg[am_f == "auto"]
mpg_m <- mtcars$mpg[am_f == "manual"]
t.test( mpg_m, mpg_a)

```

mpg by vs

```{r}
mpg_vs0 <- mtcars$mpg[vs == 0]
mpg_vs1 <- mtcars$mpg[vs == 1]
t.test( mpg_vs0, mpg_vs1)
```

### mpg by multi-level categorical variables using ANOVA

mpg and cyl 

```{r}
# Grouped Box Plot
boxplot(mpg ~ cyl, main='mpg by number of cyls',
        xlab='cyl', ylab='mpg')

# Analyse Data
aggregate(mpg ~ cyl, mtcars, mean)

# Check variances assumption
aggregate(mpg ~ cyl, mtcars, sd)

# Run ANOVA
cyl_aov <- aov(mpg ~ cyl)
summary(cyl_aov)

# Post hoc tests
TukeyHSD(cyl_aov)
```

mpg and carb

```{r}
# Grouped Box Plot
boxplot(mpg ~ carb, main='mpg by number of carbs',
        xlab='carb', ylab='mpg')
mtcars$mpg[carb == 8]
# Analyse Data
aggregate(mpg ~ carb, mtcars, mean)

# Check variances assumption
aggregate(mpg ~ carb, mtcars, sd)

# Run ANOVA
carb_aov <- aov(mpg ~ carb)
summary(carb_aov)

# Post hoc tests
TukeyHSD(carb_aov)
```

